---
title: "Appendix"
format: html
editor: visual
---

```{r, message = FALSE, echo = FALSE, warning = FALSE}
library(tidyverse)
library (performance)
library (moderndive)
library (car)
library (janitor)
library (skimr)
library (emmeans)

data_final <- read_csv("enviormental_mortality_data.csv", name_repair = make.names)

data_final <- data_final |>
 rename(PollutionExposure = PM2.5.air.pollution..mean.annual.exposure..micrograms.per.cubic.meter.)

model_final <- lm (AvgMortalityRate ~ PollutionExposure + IncomeGroup + DoctorRate , data = data_final)
```

## Appendix

### LINE assumptions

####Linearity Check 

```{r, message = FALSE, echo = FALSE, warning = FALSE}
linearity_check <- check_model(model_final, check="linearity",
                              panel=FALSE)

plot(linearity_check, data= model_final)
```

#### Normality Check 

```{r, message = FALSE, echo = FALSE, warning = FALSE}
normality_check <- check_model(model_final, check="qq",
                              panel=FALSE)

plot(normality_check, data=model_final)
```

#### Variance Check 

```{r, message = FALSE, echo = FALSE, warning = FALSE}
variance_check <- check_model(model_final, check="homogeneity",
                              panel=FALSE)

plot(variance_check, data=model_final)
```

**Linearity** 

**Independence of observations** Since our data belongs to an individual year (2017) there is very little likelihood that one observation would overlap with or influence another. Therefore, we have reason to believe that this condition is not violated.

**Normality** 

**Homogeneity of Variance**

### Regression table

```{r, message = FALSE, echo = FALSE, warning = FALSE}
summary (model_final)
```

### Contrast 1: Fixing number of doctors per 100,000

```{r, message = FALSE, echo = FALSE, warning = FALSE}

summarize(data_final, avg_pollExp = mean(PollutionExposure), std_dev = sd(PollutionExposure))

contrast1_grid <- emmeans(model_final, 
                         specs = ~  PollutionExposure + IncomeGroup + DoctorRate, 
                         at = list(PollutionExposure = c(26.04 - 20.75, 26.04, 26.04 + 20.75)))

contrast(contrast1_grid, method = "pairwise", by="PollutionExposure")
```
### Contrast 2: Fixing exposure to pollution

```{r, message = FALSE, echo = FALSE, warning = FALSE}
summarize(data_final, avg_pollExp = mean(DoctorRate), std_dev = sd(DoctorRate))

contrast2_grid <- emmeans(model_final, 
                         specs = ~  PollutionExposure + IncomeGroup + DoctorRate, 
                         at = list(DoctorRate = c(24.82 - 17.67, 24.82, 24.82 + 17.67)))

contrast(contrast2_grid, method = "pairwise", by="DoctorRate")
```


```{r, message = FALSE, echo = FALSE, warning = FALSE}
```


```{r, message = FALSE, echo = FALSE, warning = FALSE}
```


```{r, message = FALSE, echo = FALSE, warning = FALSE}
```


```{r, message = FALSE, echo = FALSE, warning = FALSE}
```


```{r, message = FALSE, echo = FALSE, warning = FALSE}
```


```{r, message = FALSE, echo = FALSE, warning = FALSE}
```


```{r, message = FALSE, echo = FALSE, warning = FALSE}
```


```{r, message = FALSE, echo = FALSE, warning = FALSE}
```


```{r, message = FALSE, echo = FALSE, warning = FALSE}
```

