---
title: "prelim_modeling"
format: html
---

```{r}
library(tidyverse)
```

```{r}
enviormental_mortality_data <- read_csv("enviormental_mortality_data.csv", name_repair = make.names)
```

```{r}
enviormental_mortality_data <- enviormental_mortality_data |>
 rename(PollutionExposureLevelAnnual = PM2.5.air.pollution..mean.annual.exposure..micrograms.per.cubic.meter.)
```


# Simple models 

```{r}
mortality_pollution <- lm(AvgMortalityRate ~ PollutionExposureLevelAnnual, data = enviormental_mortality_data)

summary(mortality_pollution)
```
- stat sig but low r^2

```{r}
mortality_doctor <- lm(AvgMortalityRate ~ DoctorRate, data = enviormental_mortality_data)

summary(mortality_doctor)
```
- stat sig but low r^2

```{r}
enviormental_mortality_data |>
  ggplot(aes(x = PollutionExposureLevelAnnual, y = AvgMortalityRate)) +
  geom_point() +
  geom_smooth(method = lm, se = FALSE, formula = y~x)
```


# Additive

```{r}
mortality_doctor_pollution_additive <- lm(AvgMortalityRate ~ DoctorRate + PollutionExposureLevelAnnual, data = enviormental_mortality_data)

summary(mortality_doctor_pollution_additive)
```
- not all coefficents are stat sig and low R^2

```{r}
mortality_income_pollution_additive <- lm(AvgMortalityRate ~ IncomeGroup + PollutionExposureLevelAnnual, data = enviormental_mortality_data)

summary(mortality_income_pollution_additive)
```
- numeric explanatory not stat sig but cat explanatory are, mediocre R^2

```{r}
mortality_income_pollution_doctor_additive <- lm(AvgMortalityRate ~ IncomeGroup + PollutionExposureLevelAnnual + DoctorRate, data = enviormental_mortality_data)

summary(mortality_income_pollution_doctor_additive)
```
- again, pollution not stat sig and mediocre R^2

```{r}
mortality_income_doctor_additive <- lm(AvgMortalityRate ~ IncomeGroup + DoctorRate, data = enviormental_mortality_data)

summary(mortality_income_doctor_additive)
```

```{r}
anova(mortality_income_doctor_additive, mortality_income_pollution_doctor_additive)
```

# Interaction 
```{r}
mortality_doctor_pollution_interaction <- lm(AvgMortalityRate ~ DoctorRate * PollutionExposureLevelAnnual, data = enviormental_mortality_data)

summary(mortality_doctor_pollution_interaction)
```
- coeffs not not stat and low R^2

```{r}
mortality_income_pollution_interaction <- lm(AvgMortalityRate ~ IncomeGroup * PollutionExposureLevelAnnual, data = enviormental_mortality_data)

summary(mortality_income_pollution_interaction)
```
- coefs mainly not stat sig and mediocre R^2

```{r}
anova(mortality_doctor_pollution_additive, mortality_doctor_pollution_interaction)
```
- says that additional complexity of the interaction is unwarranted

```{r}
mortality_income_pollution_doctor_interaction <- lm(AvgMortalityRate ~ IncomeGroup * PollutionExposureLevelAnnual * DoctorRate, data = enviormental_mortality_data)

summary(mortality_income_pollution_doctor_interaction)
```
- full interaction just for kicks, but coefficients not stat sig

```{r}
mortality_income_doctor_interaction <- lm(AvgMortalityRate ~ IncomeGroup * DoctorRate, data = enviormental_mortality_data)

summary(mortality_income_doctor_interaction)
```

```{r}
anova(mortality_income_doctor_additive, mortality_income_doctor_interaction)
```

# Various visualizations
```{r}
enviormental_mortality_data |>
  ggplot(aes(x = DoctorRate, y = AvgMortalityRate, color = IncomeGroup)) +
  geom_point()+
  geom_smooth(method = lm, formula = y~x, se=FALSE)
```

```{r}
library(moderndive)
```

```{r}
enviormental_mortality_data |>
  ggplot(aes(x = DoctorRate, y = AvgMortalityRate, color = IncomeGroup)) +
  geom_point()+
  geom_parallel_slopes(se = FALSE)
```

```{r}
enviormental_mortality_data |>
  ggplot(aes(x = IncomeGroup, y = DoctorRate)) +
  geom_point() +
  geom_jitter()
```

```{r}
enviormental_mortality_data |> 
  ggplot(aes(x = IncomeGroup, y = PollutionExposureLevelAnnual, color = IncomeGroup)) +
  geom_jitter()
```

```{r}
enviormental_mortality_data |>
  ggplot(aes(x = IncomeGroup, y = AvgMortalityRate)) +
  geom_point() +
  geom_jitter()
```





```{r}
library(performance)
```

```{r}
plot(check_outliers(mortality_income_pollution_doctor_additive))
```

