---
title: "results_analysiscode"
format: html
editor: visual
---
## loading pkgs
```{r}
library(tidyverse)
library (performance)
library (moderndive)
library (car)
library (janitor)
library (skimr)
library (emmeans)
```

```{r}
env_mortality_data <- read_csv("enviormental_mortality_data.csv", name_repair = make.names)
```

```{r}
env_mortality_data <- env_mortality_data |>
 rename(PollutionExposure = PM2.5.air.pollution..mean.annual.exposure..micrograms.per.cubic.meter.)
```

Conclusion: Model 4 is the best, the best R^2 and nested anova test passed. Adding more variables does predict AvgMortalityRate more. Suspect the Exposure being not SS from linear to multiple regression due to multicollinearity but VIFs are safe.

Questions:
1) LINE is not satisfied. Log(Y) does help the situation more, does that mean we alter all the AvgMortalityRate to log(AvgMortalityRate)? Or do we leave it as it is?

2) LINE for which other variables to check for? and what to do if each of the checks contains violations. Do facet by incomegroup?

3) Some combinations are y ~ x1 + x2 * x3. Is that something we want to explore?

Appendix?

1) for each Income Group, what does the mortality vs. Doctor Rate look like?



## 1) Mortality ~ Exposure

### 1.1) regression graph

```{r}
ggplot (data = env_mortality_data, mapping = aes ( x= PollutionExposure, y= AvgMortalityRate)) +
          geom_point ()+
          geom_smooth (method = lm, se = FALSE, formula = y~x)

```
### 1.2) model fitting 
```{r}
mortality_pollution_model <- lm (AvgMortalityRate ~ PollutionExposure, data = env_mortality_data) 

summary (mortality_pollution_model)
```
#### 1.2.1) SS - all

#### 1.2.2) R^2 - bad 

### 1.3) LINE assumptions

#### 1.3.1) Linearity - Bad 

```{r}
linearity_check_1 <- check_model(mort_pol_inc_doc_pll, check="linearity",
                              panel=FALSE)

plot(linearity_check_1, data= mort_pol_inc_doc_pll)
```
#### 1.3.2) Variance - Good
```{r}
variance_check_1 <- check_model(mort_pol_inc_doc_pll, check="homogeneity",
                              panel=FALSE)

plot(variance_check_1, data=mort_pol_inc_doc_pll)
  
```
#### 1.3.3) Normality - skewed 
```{r}
normality_check_1 <- check_model(mort_pol_inc_doc_pll, check="qq",
                              panel=FALSE)

plot(normality_check_1, data=mort_pol_inc_doc_pll)
```

#### 1.3.4) Independence - Good

#### 1.3.5) scatterplot : x vars 

```{r}
select_data <- env_mortality_data %>% 
  select (IncomeGroup,PollutionExposure, DoctorRate, AvgMortalityRate)

pairs(select_data[,-1])
```
#### 1.3.6) msc 




### 1.4) Transformation - helps

```{r}
mortality_pollution_model_log <- lm(log(AvgMortalityRate) ~ log(PollutionExposure), data = env_mortality_data)

check_model(mortality_pollution_model_log, check=c("homogeneity", "qq"))

```

### 1.5) Anova - F stat

```{r}
anova (mortality_pollution_model)
```


## 2) Mort ~ Poll + Doc 

### 2.1) regression graph


### 2.2) model fitting

```{r}
mort_pol_doc_model <- lm (AvgMortalityRate ~ PollutionExposure + DoctorRate , data = env_mortality_data) 

summary (mort_pol_doc_model)
```
#### 2.2.1) SS - not all

#### 2.2.2) R^2 - same bad

### 2.3) compare w/ int 

```{r}
mort_pol_doc_int <- lm (AvgMortalityRate ~ PollutionExposure * DoctorRate , data = env_mortality_data) 

summary (mort_pol_doc_int)
```
```{r}
anova (mort_pol_doc_model, mort_pol_doc_int)
```
#### 2.3.1) pll > int 

### 2.4) compare 2 and 1

```{r}
anova (mortality_pollution_model, mort_pol_doc_model)
```
#### 2.4.1) 1 > 2

## 3) Mort ~ Poll + Income

### 3.1) model fitting 

```{r}
mort_pol_inc_pll <- lm (AvgMortalityRate ~ PollutionExposure + IncomeGroup , data = env_mortality_data) 

summary (mort_pol_inc_pll)
```
#### 3.1.1) SS - all, exposure

#### 3.1.2) R^2 - better

### 3.2) regression graph 

```{r}
ggplot(env_mortality_data, aes(x = PollutionExposure, y = AvgMortalityRate, color = IncomeGroup)) +
  geom_point() +
  geom_parallel_slopes(method = "lm", se = FALSE)

```
### 3.3) compare w/ int

```{r}
mort_pol_inc_int <- lm (AvgMortalityRate ~ PollutionExposure * IncomeGroup , data = env_mortality_data) 

summary (mort_pol_inc_int)
```
```{r}
anova (mort_pol_inc_pll, mort_pol_inc_int)
```
- too many non-SS coefficients 

#### 3.3.1) pll>int 

### 3.4) compare 3 and 1

```{r}
anova (mortality_pollution_model, mort_pol_inc_pll)
```

#### 3.4.1) 3 > 1 

## 4) Mort ~ Poll + Inc + Doc 

#### 4.1) model fitting 

```{r}
mort_pol_inc_doc_pll <- lm (AvgMortalityRate ~ PollutionExposure + IncomeGroup + DoctorRate , data = env_mortality_data) 

summary (mort_pol_inc_doc_pll)

mort_2 <- lm (AvgMortalityRate ~ IncomeGroup + DoctorRate , data = env_mortality_data) 

summary (mort_2)

anova (mort_2, mort_pol_inc_doc_pll)
```
#### 4.1.2) SS - all, exposure 

#### 4.1.3) R^2 - better than 3

### 4.2) compare w/ int

```{r}
mort_pol_inc_doc_int <- lm (AvgMortalityRate ~ PollutionExposure * IncomeGroup * DoctorRate, data = env_mortality_data) 

summary (mort_pol_inc_doc_int)
```

```{r}
anova (mort_pol_inc_doc_pll, mort_pol_inc_doc_int)
```
#### 4.2.1) pll > int 

### 4.3) compare 4 and 3 

```{r}
anova (mort_pol_inc_pll, mort_pol_inc_doc_pll)
```
#### 4.3.1) 4 > 3

### 4.4) Compare 4 and 1

```{r}
anova (mortality_pollution_model, mort_pol_inc_doc_pll)
```
#### 4.4.1) 4 > 1 

```{r}
check_model(mort_pol_inc_doc_pll, check=c("homogeneity", "qq"))

model_performance(mort_pol_inc_doc_pll)
```

## 5) Multicollinearity: safe

- non-nested model comparison 

### 5.1) higher R^2: 4

### 5.2) VIF: low for all


```{r}
vif (mort_pol_doc_model)
```

```{r}
vif (mort_pol_inc_pll)
```

```{r}
vif (mort_pol_inc_doc_pll)
```

## 6) Informative Stats

```{r}
env_mortality_data %>% 
  skim()
```


### 6.1) Income Distribution 

```{r}
inc_dist <- env_mortality_data %>%  
  group_by (IncomeGroup) %>% 
  summarize (count = n()) %>% 
  mutate (percent = (count/81)*100)

# Global 

country_incomelevel<- read_csv("country_incomelevel.csv")%>% 
  clean_names()

inc_dist_all <- country_incomelevel %>% 
  group_by (income_group) %>% 
  summarize (count = n()) %>% 
  mutate (percent = (count/217)*100)
```

### 6.2) skimmed stats

```{r}
skimed_all <- env_mortality_data %>% 
  skim() %>% 
  filter (skim_variable == "AvgMortalityRate" |
          skim_variable == "DoctorRate"|
          skim_variable == "PollutionExposure") %>% 
  select (skim_variable, contains ("numeric"))
```

### 6.3) Poll vs. Income

```{r}
ggplot (env_mortality_data, aes (x=IncomeGroup, y = PollutionExposure))+
  geom_boxplot()

ggplot (env_mortality_data, aes (x=IncomeGroup, y = DoctorRate))+
  geom_boxplot()
```
### 6.4) Doc vs. Poll by incomegroup

```{r}
ggplot (env_mortality_data, aes (x=DoctorRate, 
                                 y = PollutionExposure, 
                                 color = IncomeGroup))+
  geom_point()+
  geom_smooth (method= lm, se = FALSE, formula = y~x)+
  theme_bw()
```

## 7) Contrasts 

### 7.1) Ref Grid 

```{r}
mort_by_inc_model <- lm (AvgMortalityRate ~ IncomeGroup + PollutionExposure, data = env_mortality_data)

mort_by_inc <- emmeans(mort_pol_inc_doc_pll, specs = ~ IncomeGroup)
mort_by_inc
```

```{r}
contrast(mort_by_inc, method = "consec")
```
### 7.2) Fixing Doc Rate 

```{r}
summarize(env_mortality_data, avg_pollExp = mean(PollutionExposure), std_dev = sd(PollutionExposure))

inc_poll_grid <- emmeans(mort_pol_inc_doc_pll, 
                         specs = ~  PollutionExposure + IncomeGroup + DoctorRate, 
                         at = list(PollutionExposure = c(26.04 - 20.75, 26.04, 26.04 + 20.75)))

contrast(inc_poll_grid, method = "pairwise", by="PollutionExposure")

```
### 7.3) Fixing Pollution exposure 

```{r}

summarize(env_mortality_data, avg_pollExp = mean(DoctorRate), std_dev = sd(DoctorRate))

inc_doc_grid <- emmeans(mort_pol_inc_doc_pll, 
                         specs = ~  PollutionExposure + IncomeGroup + DoctorRate, 
                         at = list(DoctorRate = c(24.82 - 17.67, 24.82, 24.82 + 17.67)))

contrast(inc_doc_grid, method = "pairwise", by="DoctorRate")

```









